---
title: "Credit card fraud detection"
author: "Om K, Ayush D"
date: "2024-10-11"
output: html_document
---

## Loading Libraries

Installing and loading the necessary packages.

```{r}
install.packages(c("ggplot2", "dplyr", "caret", "caTools", "ROSE", "corrplot", "glmnet"))
library(ggplot2)
library(dplyr)
library(caret)
library(caTools)
library(ROSE)  # For handling class imbalance
library(corrplot)
library(glmnet)
```

# 1. Data preprocessing

## Loading and Inspecting the Data

Reading dataset into R:

```{r}
library(readr)
creditcard <- read_csv("creditcard.csv")
View(creditcard)
```

## Checking for missing values:

```{r}
sum(is.na(creditcard))
```

# Task 2: Handling Class Imbalance

## Check the class distribution before balancing

```{r}
table(creditcard$Class)
```

## Apply Random Oversampling using the ROSE package

```{r}
set.seed(42) 
balanced_data <- ovun.sample(Class ~ ., data =
creditcard, method = "over", 
N = 2 * table(creditcard$Class)[1])$data
```

## Check the class distribution after balancing

```{r}
table(balanced_data$Class)
```

## Split the balanced data into train and test sets

```{r}

library(caTools)
set.seed(42)


split <- sample.split(balanced_data$Class, SplitRatio = 0.7)


train_data <- subset(balanced_data, split == TRUE)
test_data <- subset(balanced_data, split == FALSE)

```

## Make predictions on the test set

```{r}
predictions <- predict(model, test_data, type = "response") 
pred_class <- ifelse(predictions > 0.5, 1, 0)
```

## Evaluate the model with a confusion matrix

```{r}
library(caret) 
confusionMatrix(as.factor(pred_class), as.factor(test_data$Class))
```

# Task 3.Exploratory Data Analysis (EDA)

```{r}
summary(creditcard)
var(creditcard)
boxplot(creditcard)
plot(x=creditcard$Time, y= creditcard$Amount, xlab = "Time", ylab = "Amount", xlim = c(500, 800), ylim = c(2, 35.90),
    main = "Time vs Amount")
```

# 4. Data visualization

```{r}
# Install ggplot2 if not already installed
# install.packages("ggplot2")

# Load ggplot2 library
library(ggplot2)

# Create a scatter plot of Time vs Amount, colored by the Class (Fraud or Legitimate)
ggplot(creditcard, aes(x = Time, y = Amount, color = factor(Class))) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Credit Card Transactions: Time vs. Amount",
    x = "Time (in seconds since first transaction)",
    y = "Transaction Amount",
    color = "Class"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("0" = "blue", "1" = "red"))  # Blue for legitimate, Red for fraud

```

# Fraudulent vs Non - Fraudulent

```{r}
# Install ggplot2 if not already installed
# install.packages("ggplot2")

# Load ggplot2 library
library(ggplot2)

# Create a bar plot showing the count of fraudulent vs non-fraudulent transactions
ggplot(creditcard, aes(x = factor(Class), fill = factor(Class))) +
  geom_boxplot() +
  labs(
    title = "Fraudulent vs Non-Fraudulent Transactions",
    x = "Transaction Class",
    y = "Count of Transactions",
    fill = "Class"
  ) +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"), labels = c("Non-Fraudulent", "Fraudulent")) +
  theme_minimal()

```
